---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rlmstudio

**Disclaimer:** This is an unofficial, community-maintained R package. It is not affiliated with, endorsed by, or maintained by the creators of LM Studio.

The `rlmstudio` package provides an R interface to the [LM Studio](https://lmstudio.ai/) CLI and REST API. It allows you to download local Large Language Models (LLMs), manage server instances, load or unload models into memory, and generate text directly from your R console or scripts.

## Installation

You can install the development version of `rlmstudio` from GitHub with:

```r
# install.packages("remotes")
remotes::install_github("jmgirard/rlmstudio")
```

## Setup and Prerequisites

This package relies on the LM Studio CLI (version 0.4.0 or higher) and the new v1 REST API. If you do not have LM Studio installed or need to update your version, the package provides a convenient setup function.

```{r eval=FALSE}
library(rlmstudio)

# This will check your installation and offer to open the download page
# or perform a headless installation depending on your operating system.
install_lmstudio()
```

## Quick Start

The standard workflow involves starting the LM Studio server, loading a model into memory, chatting with it, and then cleaning up your environment.

For a deeper dive into the package architecture and a detailed explanation of how to use LM Studio with a visual GUI versus a headless background daemon, please see the `vignette("getting-started")`.

### 1. Start the Server

Before you can load models or generate text, you need to start the LM Studio backend server.

```{r eval=FALSE}
# Start the server on the default port (1234)
lms_server_start()
## ✔ LM Studio server started successfully.
```

*(Note for headless environments like remote servers or Docker containers: You must start the background daemon first using `lms_daemon_start()` before starting the server).*

### 2. Find and Load a Model

You can see which models you currently have available on your machine using `list_models()`.

```{r eval=FALSE}
# Returns a clean data frame of available models
my_models <- list_models()
my_models
##      state      type          display_name                                  key architecture size_gb
## 1 unloaded       llm          Gemma 3n E4B                  google/gemma-3n-e4b      gemma3n    5.46
## 2   loaded       llm            Gemma 3 1B                    google/gemma-3-1b  gemma3_text    0.72
## 3 unloaded       llm           Gemma 3 12B                   google/gemma-3-12b       gemma3    7.51
## 4 unloaded embedding Nomic Embed Text v1.5 text-embedding-nomic-embed-text-v1.5         <NA>    0.08
```

If you do not have a model yet, you can download one using its Hugging Face repository or LM Studio catalog identifier.

```{r eval=FALSE}
# Download a lightweight model
job_id <- lms_download("google/gemma-3-1b")
lms_download_status(job_id)
## ── Download Job: "job_02c8a1f86e"
## Status: completed
## Progress: 100% (0.72 GB / 0.72 GB)
```

Once a model is downloaded and available, load it into memory.

```{r eval=FALSE}
# Load the model
lms_load("google/gemma-3-1b")
## ✔ Model "google/gemma-3-1b" loaded and verified.
## ✔ Loading model: "google/gemma-3-1b" . . . [32.9s]
```

### 3. Chat

Use the `lms_chat()` function for quick and easy interactions.

```{r eval=FALSE}
response <- lms_chat(
  model = "google/gemma-3-1b",
  input = "Explain the difference between a data frame and a matrix in R in two sentences."
)

cat(response)
## Okay, here's the difference explained in two sentences:
##
## *   **Data frame:** A data frame is a flexible, multi-dimensional table (like a spreadsheet) in R that holds multiple variables as columns. It's the most common and versatile way to represent data.
## *   **Matrix:** A matrix is a rectangular array of numbers, often used in linear algebra and numerical computation.  It's essentially a collection of rows and columns, stored as elements in a specific format.
##
## Let me know if you'd like a more detailed explanation or want to explore examples!
```

If you need access to advanced features like Model Context Protocol (MCP) integrations, structured tool calling, or granular control over inference parameters, use the `lms_chat_advanced()` function instead.

### 4. Clean Up

When you are finished, it is good practice to unload the model from memory and shut down the local server to free up system resources.

```{r eval=FALSE}
# Unload the specific model
lms_unload("google/gemma-3-1b")
## ✔ Model "google/gemma-3-1b" unloaded successfully.
## ✔ Unloading model: "google/gemma-3-1b"... [374ms]

# Stop the local server
lms_server_stop()
## ✔ LM Studio server stopped successfully.
```
