---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lms

<!-- badges: start -->
<!-- badges: end -->

The `lms` package is a wrapper for the [LM Studio](https://lmstudio.ai) [Command Line Interface](https://lmstudio.ai/docs/cli), allowing you to access and control the LM Studio software directly from R.

## Installation

You can install the development version of `lms` from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("pak")
pak::pak("jmgirard/lms")
```

## Setup

Before using the package, you need to ensure the LM Studio CLI is installed and updated to version 0.4.0 or higher. You can check your setup or launch the installation process using:

```{r, eval=FALSE}
library(lms)

# Check installation status
has_lms()

# Help install or update the CLI
lms_setup(method = "browser") # or "headless" for Linux/servers
```

## Example Usage

### Managing Models

You can list downloaded models, search for new ones, and load them into memory. If you want to process the model lists programmatically, you can return them as JSON-parsed data frames.

```{r, eval=FALSE}
# List all local LLMs as a data frame
my_models <- model_ls(llm = TRUE, json = TRUE)
my_models
```

To load and chat with a model interactively or programmatically:

```{r, eval=FALSE}
# Load a model into memory
model_load("llama-3.1-8b", context_length = 2048)

# Send a prompt and capture the response
response <- model_chat(
  model = "llama-3.1-8b",
  prompt = "What is the capital of France?",
  capture = TRUE
)
response

# Unload when finished
model_unload("llama-3.1-8b")
```

### Server and Daemon Control

For headless environments or automated scripts, you can control the background daemon and local API server.

```{r, eval=FALSE}
# Start the background daemon
daemon_up()

# Start the local server on port 1234
server_start(port = 1234)

# Check server status
server_status(json = TRUE)

# Stop the server and daemon
server_stop()
daemon_down()
```
