---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rlmstudio

**Disclaimer:** This is an unofficial, community-maintained R package. It is not affiliated with, endorsed by, or maintained by the creators of LM Studio.

The `rlmstudio` package provides an R interface to the [LM Studio](https://lmstudio.ai/) CLI and REST API. It allows you to download local Large Language Models (LLMs), manage server instances, load or unload models into memory, and generate text directly from your R console or scripts.

## Installation

You can install the development version of `rlmstudio` from GitHub with:

```r
# install.packages("remotes")
remotes::install_github("jmgirard/rlmstudio")
```

## Setup and Prerequisites

This package relies on the LM Studio CLI (version 0.4.0 or higher) and the new v1 REST API. If you do not have LM Studio installed or need to update your version, the package provides a convenient setup function.

```{r eval=FALSE}
library(rlmstudio)

# This will check your installation and offer to open the download page
# or perform a headless installation depending on your operating system.
install_lmstudio()
```

## Quick Start

The standard workflow involves starting the LM Studio server, loading a model into memory, chatting with it, and then cleaning up your environment.

For a deeper dive into the package architecture and a detailed explanation of how to use LM Studio with a visual GUI versus a headless background daemon, please see the `vignette("getting-started")`.

### 1. Start the Server

Before you can load models or generate text, you need to start the LM Studio backend server.

```{r eval=FALSE}
# Start the server on the default port (1234)
lms_server_start()
```

*(Note for headless environments like remote servers or Docker containers: You must start the background daemon first using `lms_daemon_start()` before starting the server).*

### 2. Find and Load a Model

You can see which models you currently have available on your machine using `list_models()`.

```{r eval=FALSE}
# Returns a clean data frame of available models
my_models <- list_models()
my_models
```

If you do not have a model yet, you can download one using its Hugging Face repository or LM Studio catalog identifier.

```{r eval=FALSE}
# Download a lightweight model
job_id <- lms_download("hf.co/bartowski/Llama-3.2-1B-Instruct-GGUF")
lms_download_status(job_id)
```

Once a model is downloaded and available, load it into memory.

```{r eval=FALSE}
# Load the model
lms_load("hf.co/bartowski/Llama-3.2-1B-Instruct-GGUF")
```

### 3. Chat

Use the `lms_chat()` function for quick and easy interactions.

```{r eval=FALSE}
response <- lms_chat(
  model = "hf.co/bartowski/Llama-3.2-1B-Instruct-GGUF",
  input = "Explain the difference between a data frame and a matrix in R in two sentences."
)

cat(response)
```

If you need access to advanced features like Model Context Protocol (MCP) integrations, structured tool calling, or granular control over inference parameters, use the `lms_chat_advanced()` function instead.

### 4. Clean Up

When you are finished, it is good practice to unload the model from memory and shut down the local server to free up system resources.

```{r eval=FALSE}
# Unload the specific model
lms_unload("hf.co/bartowski/Llama-3.2-1B-Instruct-GGUF")

# Stop the local server
lms_server_stop()
```
