---
title: "Getting Started with lms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette provides a step-by-step guide to using the `lms` package to control LM Studio from R. The workflow below focuses on using the local API server for programmatic chat interactions.

## Prerequisites

Before using the R package, you need to set up the LM Studio software on your machine.

### 1. Install LM Studio GUI
Visit [lmstudio.ai](https://lmstudio.ai) to download and install the version appropriate for your operating system. The GUI provides a user-friendly interface for managing models and local server settings.

### 2. Browse and Download Models
Open the LM Studio application. Use the search/browser tab to find a model. For this tutorial, search for a small, efficient model like **Gemma 3 1B** (~1GB) or **Gemma 3n** (~6GB).

* Small models are recommended for initial testing as they download and load into memory quickly.
* Once found, click the **Download** button for your preferred version.

## R Package Workflow

### 3. Install and Load the Package
You can install the development version of `lms` from GitHub.

```{r}
#| eval: false
# install.packages("pak")
pak::pak("jmgirard/lms")
library(lms)
```

### 4. Start the API Server
The LM Studio local server allows you to interact with models via HTTP API calls. Run the following command to launch the server:

```{r}
#| eval: false
server_start()
## ✔ LM Studio server started successfully.
```

* By default, the server runs on port 1234.
* This is necessary to enable functions like `api_chat()` to communicate with the backend.

### 5. Load the Model into Memory
Before you can chat, the model must be loaded into your system's memory (RAM or VRAM). Use the identifier or path for the model you downloaded:

```{r}
#| eval: false
model_load("google/gemma-3-1b")
## ✔ Model loaded successfully.
```

* Loading the model explicitly ensures it is ready to process requests immediately.

### 6. Interaction via api_chat
The `api_chat()` function is a convenience wrapper for simple, one-turn interactions.

#### User Prompts
To send a basic message and get a simplified text response, use the `prompt` argument:

```{r}
#| eval: false
response <- api_chat(prompt = "say hello")
print(response)
```

#### System Prompts
You can influence the model's behavior or persona using a `system_prompt`:

```{r}
#| eval: false
rhyme_response <- api_chat(
  prompt = "say hello",
  system_prompt = "answer in rhymes"
)
print(rhyme_response)
```

### 7. Adjusting Context Length
If you are working with long documents or extensive conversation history, you may need to increase the `context_length`. This is the number of tokens the model considers during processing. You set this during the loading phase:

```{r}
#| eval: false
# Reload with a larger context window
model_load("google/gemma-3n-e4b", context_length = 4096)
```

## Cleanup

It is good practice to release system resources when you are finished.

### 8. Unload and Stop
First, remove the model from memory, then stop the local server.

```{r}
#| eval: false
# Unload the specific model
model_unload("google/gemma-3n-e4b")

# Stop the API server
server_stop()
```

* `model_unload()` frees up your RAM/GPU memory for other tasks.
* `server_stop()` shuts down the background process managing the API.
