---
title: "Getting Started with rlmstudio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rlmstudio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `rlmstudio` package bridges the gap between R and local Large Language Models (LLMs) by wrapping the LM Studio CLI and its v1 REST API. This vignette will guide you through the two primary ways to run LM Studio and show you how to manage and chat with your models.

## Choosing Your Workflow: GUI vs. Headless

LM Studio can run its local server in two different ways. The R package works identically with both, but how you start the server depends on your preference and operating system.

### 1. The GUI Workflow (Desktop Application)
* **Best for:** Visual users, desktop environments (macOS, Windows, and Linux desktops), and those who like to manually manage their models.
* **How it works:** You open the LM Studio desktop application, navigate to the Developer/Local Server tab, and click "Start". You do not need to use the startup functions in R.
* **The R Code:** You simply open R and immediately start running commands like `lms_load()` and `lms_chat()`. The R package communicates directly with the server you started in the visual app.

### 2. The Headless Workflow (CLI Daemon)
* **Best for:** Automated R scripts, remote servers, Docker containers, and users who want R to manage the entire process from start to finish.
* **How it works:** The GUI remains completely closed. Instead, you use R to boot up the LM Studio background engine (the daemon) and the local server.
* **The R Code:** You must run `lms_daemon_start()` followed by `lms_server_start()` before you can load models or chat. When your script finishes, you clean up with `lms_server_stop()` and `lms_daemon_stop()`.
* **Important Note:** If you try to run `lms_daemon_start()` while the desktop GUI is open, you will encounter conflicts. Pick one lane and stick to it for your session.

## Step-by-Step Guide

### 1. Start the Environment

First, load the package and spin up the (daemon and) server.

```{r setup, eval=FALSE}
library(rlmstudio)

# If headless, start the background daemon (if not, start the GUI)
lms_daemon_start()
## ✔ LM Studio daemon started in the background.

# Start the local server
lms_server_start()
## ✔ LM Studio server started successfully.
```

### 2. Finding and Managing Models

You can check which models you have downloaded (e.g., through the GUI) and whether they are currently loaded into memory using `list_models()`.

```{r list-models, eval=FALSE}
# View all models
models <- list_models()

# You can use standard R tools to filter this list
# For example, finding all unloaded llm models:
unloaded_llms <- models |>
  subset(type == "llm" & state == "unloaded")
unloaded_llms
##      state type display_name                 key architecture size_gb
## 1 unloaded  llm   Gemma 3 1B   google/gemma-3-1b  gemma3_text    0.72
## 2 unloaded  llm   Gemma 3 4B   google/gemma-3-4b       gemma3    2.83
## 3 unloaded  llm Gemma 3n E4B google/gemma-3n-e4b      gemma3n    5.46
## 4 unloaded  llm  Gemma 3 12B  google/gemma-3-12b       gemma3    7.51
```

If you need a new model, you can download it through the package (or the GUI). Package downloads happen asynchronously in the background.

```{r download, eval=FALSE}
job_id <- lms_download("qwen/qwen3-4b-2507")

# Check the progress
lms_download_status(job_id)
## ── Download Job: "job_493c7c9ded"
## Status: completed
## Progress: 100% (2.12 GB / 2.12 GB)
```

### 3. Loading Models

Before you can chat with a model, you must load it into system memory (RAM or VRAM). You can pass specific hardware parameters during this step if needed.

```{r load, eval=FALSE}
# Standard load
lms_load("qwen/qwen3-4b-2507")
## ✔ Model "qwen/qwen3-4b-2507" loaded and verified. [27.6s]

# Advanced load with specific parameters
lms_load(
  model = "qwen/qwen3-4b-2507",
  context_length = 4096,
  flash_attention = TRUE
)
## ✔ Model "qwen/qwen3-4b-2507" loaded and verified. [7.5s]
```

### 4. Chatting

The `lms_chat()` function is designed for simple, single-turn prompts.

```{r simple-chat, eval=FALSE}
response <- lms_chat(
  model = "qwen/qwen3-4b-2507",
  input = "Say hello!",
  system_prompt = "Answer in rhymes."
)

cat(response)
## Hello there, friend! 🌟
## With a smile, I greet you—what a delight!
## Waves of joy and warmth, I’m here to stay,
## Spreading light through the quiet day.
##
## So say hello back, and let the fun begin—
## I’m ready to chat, or just be your friend.
## No need for books, or charts, or codes,
## Just a spark, a thought, or a simple "how are you?" — I’m at your door!
##
## So here’s to dreams, and giggles, and rhyme,
## Hello, world—this is how we climb! 🌈✨
```

For more complex use cases (like streaming, adjusting temperatures, or utilizing Model Context Protocol integrations), use `lms_chat_advanced()`. Setting `simplify = FALSE` will return the raw, unparsed API list response.

```{r advanced-chat, eval=FALSE}
advanced_response <- lms_chat_advanced(
  model = "qwen/qwen3-4b-2507",
  input = "Explain variance.",
  temperature = 0.2,
  max_output_tokens = 150,
  simplify = FALSE
)

# Access the generated text from the structured list
advanced_response$output[[1]]$content
## [1] "**Variance** is a statistical measure that quantifies how spread out a set of data points is from their mean (average value). It tells us the average of the squared differences between each data point and the mean.\n\n### Formula for Variance:\nFor a population:\n\\[\n\\sigma^2 = \\frac{1}{N} \\sum_{i=1}^{N} (x_i - \\mu)^2\n\\]\nFor a sample:\n\\[\ns^2 = \\frac{1}{n-1} \\sum_{i=1}^{n} (x_i - \\bar{x})^2\n\\]\n\nWhere:\n- \\( \\sigma^2 \\) = population variance  \n- \\( s^"
```

### 5. Teardown

To free up memory, unload your models when you are done. If you are running the headless setup, shut down the server and daemon.

```{r teardown, eval=FALSE}
# Unload the model
lms_unload("qwen/qwen3-4b-2507")
## ✔ Model "qwen/qwen3-4b-2507" unloaded successfully. [391ms]

# Stop the server
lms_server_stop()
## ✔ LM Studio server stopped successfully.

# If headless, stop the daemon (if not, just close the GUI)
lms_daemon_stop()
```
