---
title: "Headless Configuration and Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Headless Configuration and Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(rlmstudio)

# Two gates for the chunks below. The first says the CLI is on this machine.
# The second says the REST API answered. It stays FALSE until the server has
# been started and asked.
lms_installed <- has_lms()
lms_ready <- FALSE

model <- "google/gemma-3-1b"

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `rlmstudio` package provides robust support for running LM Studio in completely headless environments. This is ideal for Linux servers, Docker containers, remote cloud instances, and automated CI/CD pipelines where a visual desktop application is unavailable or inconvenient.

To operate without a GUI, LM Studio relies on a background process called the `llmster` daemon. This vignette will walk you through managing the daemon, starting the local server, and fully automating your local LLM workflows.

## Setup and Installation

If you are setting up a fresh remote server, you can use the package to download and install the LM Studio CLI automatically via the terminal. Run `install_lmstudio(method = "headless")` in your console to execute the automated installation script.

```{r check-status}
# Verify the CLI is available before proceeding
has_lms()
```

## Step-by-Step Guide

### 1. Start the Background Daemon

Unlike the desktop version where opening the app initializes the backend engine, a headless environment requires you to start the engine manually. You must start the `llmster` daemon before attempting to load models or start the API server.

```{r start-daemon, eval=lms_installed}
# Start the headless engine in the background
lms_daemon_start()
```

### 2. Start the Local Server

With the daemon running, you can now spin up the REST API server to accept HTTP requests. The CLI returns before the REST API answers, so `lms_server_start()` keeps asking the REST API whether it is ready and returns once it answers. A headless box is often the slow one, so allow it more than the default 10 seconds.

```{r start-server, eval=lms_installed}
# Start the local server on the default port, and allow 30 seconds for it
lms_server_start(wait = 30)
```

A wait that runs out raises a warning and returns, so on a headless box you
still check that the server answers before you call it. `lms_server_ready()`
asks the host for a model list. It reports `FALSE` for a port held by another
process, for a server that is still coming up, and for a server that turned
your token away.

```{r check-ready, eval=lms_installed}
lms_ready <- lms_server_ready()
lms_ready
```

### 3. Finding and Managing Models

Because you do not have the GUI's visual search tool, you will need to know the Hugging Face repository or the LM Studio catalog identifier for the model you want to use.

```{r download, eval=lms_ready}
# Download a model using its identifier
job_id <- lms_download("qwen/qwen3-4b-2507")
lms_download_status(job_id)
```

```{r wait, echo=FALSE, eval=lms_ready}
if (job_id != "already_downloaded") {
  repeat {
    res <- lms_download_status(job_id)
    if (res$status %in% c("completed", "failed", "error")) {
      break
    }
    Sys.sleep(3)
  }
  if (res$status != "completed") {
    cli::cli_abort("Model download failed.")
  }
}
```

```{r list, eval=lms_ready}
# View all downloaded models
models <- list_models()

# Filter for unloaded text models
unloaded_llms <- models |>
  subset(type == "llm" & state == "unloaded")
unloaded_llms
```

### 4. Loading Models

Allocate the model to your system's memory (RAM/VRAM) so it is ready for inference.

```{r load, eval=lms_ready}
# Load the model
lms_load(model, flash_attention = TRUE)
```

### 5. Chatting

Interact with the model exactly as you would in a desktop environment.

```{r chat, eval=lms_ready}
response <- lms_chat(
  model = model,
  input = "Provide just the str_extract() pattern to match all text after the third comma.",
  system_prompt = "You are an expert R programmer familiar with the tidyverse."
)

cat(response)
```

### 6. Teardown and Cleanup

In a headless environment, managing your system resources is critical. When your script finishes, you should explicitly tear down the entire stack to free up memory and stop background processes.

```{r unload, eval=lms_ready}
# 1. Unload the model from memory
lms_unload(model)
```

Stopping the server and the daemon does not go through the REST API, so it
runs whenever the CLI is here. That way a stack this vignette started is torn
down even if the readiness check said no.

```{r stop-stack, eval=lms_installed}
# 2. Stop the API server
lms_server_stop()

# 3. Stop the background daemon
lms_daemon_stop()
```

## Bonus: Pipeline Automation

If you are writing a script that just needs to run a quick job and exit, managing the daemon state manually can be tedious. The `with_lms_daemon()` wrapper handles the setup and guaranteed teardown of the background engine automatically.

This block starts its own server, so it asks again whether that server answers.
The value measured earlier belongs to the server the teardown above has already
stopped.

```{r with-daemon, eval=lms_installed}
# The daemon will start, the code will run, and the daemon will stop on exit.
results <- with_lms_daemon({
  lms_server_start()

  res <- NULL
  if (lms_server_ready()) {
    lms_load(model)
    res <- lms_chat(model, "Is the daemon running?")
  }

  lms_server_stop()
  res
})
```
