---
title: "Text Analysis with LM Studio in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Text Analysis with LM Studio in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(rlmstudio)

# Gates for the chunks below. The first says the CLI is on this machine.
# The second says the REST API answered. The last two say that the chat
# model and the embedding model are on disk. The last three stay FALSE
# until the server has been started and asked.
lms_installed <- has_lms()
lms_ready <- FALSE
chat_ready <- FALSE
embed_ready <- FALSE

model <- "google/gemma-3-1b"
embed_model <- "text-embedding-nomic-embed-text-v1.5"

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to use a local model to analyze a set of texts. It
sends a batch of prompts and gets the replies back as a data frame. It asks
for replies in a fixed shape, so that each field becomes a column. It turns
the token probabilities of a one-digit rating into a score. It also turns
texts into embedding vectors. The vignette `vignette("getting-started")`
covers the steps that come first: installing LM Studio, starting the
server, and downloading `google/gemma-3-1b`.

## Start the server

```{r server-before, include=FALSE}
# Whether the server ran before this vignette. The teardown stops only a
# server that this vignette started.
server_was_running <- FALSE
if (lms_installed) {
  status <- suppressWarnings(lms_server_status(json = TRUE))
  server_was_running <- is.list(status) && isTRUE(status[["running"]])
}
```

```{r start-server, eval=lms_installed}
lms_server_start(wait = 30)
#> ✔ LM Studio server started successfully on the default port.
lms_ready <- lms_server_ready()
lms_ready
#> [1] TRUE
```

```{r models-before, include=FALSE}
# The ids of the instances that were loaded before this vignette. The
# teardown unloads only the instances that this vignette loaded.
loaded_ids <- function() {
  loaded <- list_models(loaded = TRUE, detailed = TRUE, quiet = TRUE)
  ids <- lapply(loaded$loaded_instances, function(x) x[["id"]])
  as.character(unlist(ids))
}
ids_before <- character()
if (lms_ready) {
  on_disk <- list_models(quiet = TRUE)$key
  chat_ready <- model %in% on_disk
  embed_ready <- embed_model %in% on_disk
  ids_before <- loaded_ids()
}
```

## Quiet output

`lms_load()` prints messages while it loads a model, and `lms_chat_batch()`
shows a progress bar while a batch runs. Set the `rlmstudio.quiet` option to
`TRUE` to hide both. The warning that a batch gives about failed inputs
still shows. This vignette sets the option here and sets it back before the
teardown.

```{r quiet}
old_options <- options(rlmstudio.quiet = TRUE)
```

```{r load, eval=chat_ready}
lms_load(model)
```

## A batch of prompts as a data frame

`lms_chat_batch()` sends each input as its own request. With
`format = "data.frame"`, it returns one row per input. The row holds the
input, the reply text in `output`, the id of the reply, and three token
counts.

```{r batch, eval=chat_ready}
reviews <- c(
  "The food was great but the service was slow.",
  "Terrible. Never again.",
  "Best pizza in town, and the staff were friendly."
)
summaries <- lms_chat_batch(
  model,
  reviews,
  system_prompt = "Summarize the review in five words or fewer.",
  format = "data.frame",
  temperature = 0
)
names(summaries)
#> [1] "input"                   "output"
#> [3] "response_id"             "input_tokens"
#> [5] "total_output_tokens"     "reasoning_output_tokens"
summaries[, c("input", "output")]
#>                                              input
#> 1     The food was great but the service was slow.
#> 2                           Terrible. Never again.
#> 3 Best pizza in town, and the staff were friendly.
#>                               output
#> 1       Excellent, but slow service.
#> 2 **A disappointing experience.** \n
#> 3                   Excellent pizza!
```

## Structured output as columns

A `schema` goes to the server as a JSON schema that the reply must match. On
`api_type = "openai"`, a data-frame batch with an object schema adds one
column per top-level property of the schema. A `"string"` property gives a
character column, and an `"integer"` property gives an integer column.

```{r schema, eval=chat_ready}
schema <- list(
  type = "object",
  properties = list(
    sentiment = list(
      type = "string",
      enum = list("positive", "negative", "mixed")
    ),
    stars = list(type = "integer")
  ),
  required = list("sentiment", "stars")
)
rated <- lms_chat_batch(
  model,
  reviews,
  system_prompt = paste(
    "Give the sentiment of the review, and rate it from 1 to 5 stars,",
    "where 1 is the worst and 5 is the best."
  ),
  format = "data.frame",
  api_type = "openai",
  schema = schema,
  temperature = 0
)
rated[, c("input", "sentiment", "stars")]
#>                                              input sentiment stars
#> 1     The food was great but the service was slow.     mixed     4
#> 2                           Terrible. Never again.  negative     5
#> 3 Best pizza in town, and the staff were friendly.  positive     5
```

In this output, the model gave 5 stars to the negative review. A model as
small as `google/gemma-3-1b` makes mistakes of this kind, so read its
replies before you rely on them.

## A rating scored from token probabilities

A model that answers with one digit picks that digit from several
candidates. With `logprobs = TRUE`, `lms_chat()` returns an object that
holds the reply text and a data frame of log probabilities, if the reply
carries them. The data frame lists the candidate tokens at each step of the
reply. The `top_logprobs` and `temperature` fields go into the request body.

```{r logprobs, eval=chat_ready}
rating <- lms_chat(
  model = model,
  input = paste(
    "Rate how positive this review is from 1 to 5.",
    "Answer with one digit only.",
    "Review: The food was great but the service was slow."
  ),
  logprobs = TRUE,
  top_logprobs = 10,
  temperature = 0
)
rating$text
#> [1] "3\n"
head(rating$logprobs, 3)
#>   step_token step_logprob candidate_token candidate_logprob
#> 1          3     -0.34375               3         -0.343750
#> 2          3     -0.34375               4         -1.546875
#> 3          3     -0.34375               5         -2.750000
```

`lms_score_expected()` reads the rows whose `step_token` equals that of the
first row. In the reply above, those rows hold the candidates for the digit.
It keeps the candidates whose token is a number in `scale`, and it scales
their probabilities to sum to 1. It returns the expected value, the weighted
standard deviation, the entropy in bits, and the probability of each label.

```{r score, eval=chat_ready}
score <- lms_score_expected(rating$logprobs, scale = 1:5)
score$expected_value
#> [1] 3.342552
score$probabilities
#>   label         prob
#> 1     3 0.7176032364
#> 2     4 0.2154635645
#> 3     5 0.0646938939
#> 4     2 0.0021793830
#> 5     1 0.0000599222
```

## Embeddings

`lms_embed()` turns each text into a vector of numbers. It returns a matrix
with one row per input text, in the order given. LM Studio comes with the
embedding model `text-embedding-nomic-embed-text-v1.5`.

```{r embed, eval=embed_ready}
lms_load(embed_model)
texts <- c(reviews, "The stock market fell sharply today.")
vectors <- lms_embed(embed_model, texts)
dim(vectors)
#> [1]   4 768

# The cosine similarity of each pair of texts
unit <- vectors / sqrt(rowSums(vectors^2))
round(unit %*% t(unit), 2)
#>      [,1] [,2] [,3] [,4]
#> [1,] 1.00 0.37 0.56 0.47
#> [2,] 0.37 1.00 0.36 0.44
#> [3,] 0.56 0.36 1.00 0.34
#> [4,] 0.47 0.44 0.34 1.00
```

## Loaded instances

`list_instances()` returns one row per loaded model instance. The columns
hold the instance id, the model key, the model type, the display name, and
the load configuration of the instance.

```{r instances, eval=lms_ready}
instances <- list_instances()
instances[, c("id", "type", "context_length")]
#>                                     id      type context_length
#> 1                    google/gemma-3-1b       llm           8192
#> 2 text-embedding-nomic-embed-text-v1.5 embedding           2048
```

## Teardown

```{r quiet-reset}
# Set the option back to the value it had before this vignette
options(old_options)
```

`lms_unload_all()` unloads every loaded model instance. When this vignette
is built with the server stopped and no model loaded, it calls
`lms_unload_all()` and then stops the server. When it is built with the
server running and `google/gemma-3-1b` loaded, it unloads only the
instances that it loaded, and it leaves the server running.

```{r ids-after, include=FALSE}
ids_after <- if (lms_ready) loaded_ids() else character()
```

```{r unload-all, eval=lms_ready && length(ids_before) == 0}
# The build found no model loaded, so it unloads every instance
lms_unload_all()
#> ℹ Found 2 loaded model instances. Unloading now...
#> ℹ Unloading model: "google/gemma-3-1b"...
#> ✔ Model "google/gemma-3-1b" unloaded successfully. [564ms]
#>
#> ℹ Unloading model: "text-embedding-nomic-embed-text-v1.5"...
#> ✔ Model "text-embedding-nomic-embed-text-v1.5" unloaded successfully. [21ms]
#>
#> ✔ All models unloaded successfully.
```

```{r unload-new, eval=lms_ready && length(ids_before) > 0}
# The build found a model loaded, so it unloads only what it loaded
for (id in setdiff(ids_after, ids_before)) {
  lms_unload(id)
}
#> ℹ Unloading model: "text-embedding-nomic-embed-text-v1.5"...
#> ✔ Model "text-embedding-nomic-embed-text-v1.5" unloaded successfully. [27ms]
```

```{r stop-server, eval=lms_installed && !server_was_running}
lms_server_stop()
#> ✔ LM Studio server stopped successfully.
```
